\bf Adapting the Chumbley Score to match striae on Land Engraved Areas of bullets

### 2018-08-19 doc/paper.tex
resolution of images

  \author{
        Ganesh Krishnan,\\ %\\
    Master of Science (M.Sc.)\\
    % \\
     and \\     Prof.~Heike Hofmann,\\ %\\
    Doctor of Philosophy (Ph.D.)\\
    % \\
      }
    \maketitle
  \bigskip
  \begin{center}
  {
  \textbf{Affiliation(s)}\\
  Department of Statistics, Iowa State University and CSAFE (Center for
  Statistics and Applications in Forensic Evidence)\\
  \bigskip
  \textbf{Source of funding}\\
  The authors gratefully acknowledge funding from CSAFE through
  cooperative agreement \# 70NANB15H176 between NIST and Iowa State
  University.\\
  \bigskip
  \textbf{Acknowledgement}\\
  We would like to thank the efforts of CSAFE and Roy J Carver Microscopy
  lab in scanning the Hamby-44 set and providing the scans to us. We would
  also like to thank NIST for providing the Hamby-44 bullets to us for
  scanning.\\
  \bigskip
  \textbf{Information on presentation of work}\\
  This work was presented in the form of a poster in the AFTE 2018 49th
  Annual Training Seminar, held from June 3rd to June 8th, 2018 in
  Charleston, West Virginia
  }
  \end{center}

@@

Methods for matching {\textcolor{blue}{striated}} marks for a variety of
tools have been studied in the literature (see
\autoref{tab:toolmarks-ER} for an overview): in \cite{manytoolmarks1}
and \cite{chumbley} digitized screwdriver marks have been analyzed using
a profilometer; in \cite{manytoolmarks2} 3D marks from screwdriver,
tongue and groove pliers captured using a confocal microscope have been
investigated; digitized marks from slip-joint pliers generated by a
surface profilometer have been investigated in \cite{afte-chumbley}.

{\textcolor{blue}{changed}} These digitized markings allow the use of
statistical methods to quantify the scientific mechanism of comparing
markings and serve as basis for an error rate calculation. In
\cite{manytoolmarks2} a relative distance metric is defined and used as
similarity measure between two toolmarks. This approach is expanded in
\citet{manytoolmarks1}: a set of small segments in the markings of two
toolmarks are extracted and similarity is compared using a maximum
Pearson correlation coefficient. The Chumbley scoring method, first
introduced in \cite{chumbley}, uses a similar but more extensive
framework based on a Mann-Whitney U test of the resulting correlation
coefficients. This approach is non-deterministic, because segments are
chosen randomly. In \cite{hadler} the score is fixed to be deterministic
for each pair of toolmarks by choosing segments for comparison
systematically. This approach also ensures independence between segments
of striae.

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{rlrrr}
  \hline
Research paper & Method & Data Source & False Positives & False Negatives\\
  \hline
 \textbf{Faden et al. (2007)} \cite{manytoolmarks1} & Maximum Pearson & Screwdrivers & & \\ & Correlation &  & - & - \\ \hline
 \textbf{Chumbley et al. (2010)} \cite{chumbley} & Randomized & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 2.3\% & 8.9\%  \\ \hline
  \textbf{Grieve et al. (2014)}
 \cite{afte-chumbley} & Randomized & Slip-joint  &  & \\ & Chumbley Score &  & - & - \\ \hline 
  \textbf{Hadler \& Morris (2017)} \cite{hadler} & Deterministic & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 0\% & 6\%  \\ \hline
  \textbf{Bachrach et al. (2010)} \cite{manytoolmarks2} & Similarity Measure & Screwdrivers & & \\ (Different Surfaces-same angle) & Relative Distance Metric & & 5.9\% &  9.4\% \\ (Same Surfaces-same angle)&  & & 0.22\% &  0\% \\
   \hline
\end{tabular}}
\caption{\label{tab:toolmarks-ER} Error Rates in same-source {\textcolor{blue}{striated}} toolmark analysis reported in the literature. The top four reported papers use some variation of the Chumbley-score method.}
\end{table}

@@

\begin{table}[hbtp]
\centering
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{rrrlrrl}
  \hline
  & \multicolumn{3}{l}{\bf CS1} & \multicolumn{3}{l}{\bf CS2} \\
$w_o$ & \bf FPR & \bf FNR & \bf AUC (95\% C.I.) & \bf FPR & \bf FNR & \bf AUC (95\% C.I.) \\ 
  \hline
90 & 0.068 & 0.330 & 0.850 (0.835, 0.865) & 0.075 & 0.242 & 0.877 (0.863, 0.890) \\ 
  120 & 0.065 & 0.309 & 0.864 (0.850, 0.878) & 0.072 & 0.236 & 0.890 (0.877, 0.903) \\ 
  130 & 0.065 & 0.317 & 0.863 (0.850, 0.877) & 0.072 & 0.224 & 0.898 (0.886, 0.910) \\ 
  140 & 0.065 & 0.304 & 0.873 (0.859, 0.886) & 0.071 & 0.229 & 0.902 (0.891, 0.914) \\ 
  150 & 0.067 & 0.326 & 0.863 (0.850, 0.877) & 0.072 & 0.225 & 0.904 (0.892, 0.916) \\ 
  180 & 0.061 & 0.326 & 0.877 (0.864, 0.890) & 0.070 & 0.229 & 0.907 (0.896, 0.919) \\ 
  210 & 0.064 & 0.342 & 0.865 (0.851, 0.878) & 0.066 & 0.231 & 0.906 (0.895, 0.918) \\ 
   \hline
\end{tabular}
\end{adjustbox}
\caption{\label{tab:aucs} Overview of results as shown in Figures \ref{fig:roc} and \ref{fig:eer}. FPR is the observed rate of false positives (for a fixed level $\alpha = 0.05$), FNR is the false negative rate. Area under the curve (AUC) is shown with confidence intervals. }
\end{table}

@@

\autoref{fig:type1} shows observed type I error rates (FPR) across a
range of optimization windows \(w_o\). Generally, observed type I errors
are higher than expected. Method CS1 shows in this instance slightly
better performance than method CS2, but for both an increase in the size
of the optimization window leads to a decrease in the observed type I
error rates.

@@

The high-resolution scans of Hamby set 44 are capturing images at a
resolution of \(0.645 \mu m\) per pixel. On average, land engraved areas
are 3000 pixels in length. A coarseness of \(c = 0.125\) seemed to be
sufficient in removing any bullet curvature. Both methods have a failed
test rate of less than 0.006, indicating, again, that the larger number
of pixels alleviates the problem of test failures. \autoref{fig:h44}
shows the resulting EER and AUC for methods CS1 and CS2 based on two
sizes of validation windows (\(w_v \in 75, 125\)) and optimization
window sizes around 300 pixels (10 percent of the average length). Both
methods show an increase in performance around \(w_o = 300\) pixels. CS2
out-performs CS1 in all scenarios, but the difference is not significant
(using DeLong's confidence intervals). Interestingly, the overall
performance of both CS1 and CS2 is a lot lower for the high resolution
version of Hamby-44 than for the lower resolution scans. The area under
the curve overall is significantly lower for the high resolution scans
than for the previous set of scans. Partly, this might be due to the
particular choice of the parameters, partly the higher-resolution scans
might be picking up on real differences between the lands that the
lower-resolution scans fail to detect.

@@

\begin{table}[hbtp]
\centering
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{rrrlrrl}
  \hline
  & \multicolumn{3}{l}{\bf CS1} & \multicolumn{3}{l}{\bf CS2} \\
$w_o$ & \bf FPR & \bf FNR & \bf AUC (95\% C.I.) & \bf FPR & \bf FNR & \bf AUC (95\% C.I.) \\  
  \hline
%90 & 0.077 & 0.546 & 0.680 (0.633, 0.727) & 0.056 & 0.478 & 0.729 (0.684, 0.774) \\
%  120 & 0.082 & 0.485 & 0.736 (0.693, 0.779) & 0.056 & 0.452 & 0.763 (0.720, 0.806) \\
%  150 & 0.080 & 0.493 & 0.730 (0.685, 0.775) & 0.055 & 0.422 & 0.774 (0.733, 0.816) \\
%  180 & 0.087 & 0.458 & 0.773 (0.734, 0.813) & 0.055 & 0.435 & 0.773 (0.730, 0.816) \\
  210 & 0.086 & 0.413 & 0.780 (0.740, 0.821) & 0.055 & 0.387 & 0.803 (0.763, 0.843) \\ 
  240 & 0.085 & 0.427 & 0.784 (0.744, 0.824) & 0.055 & 0.365 & 0.801 (0.760, 0.841) \\ 
  270 & 0.082 & 0.415 & 0.782 (0.741, 0.823) & 0.048 & 0.374 & 0.808 (0.769, 0.848) \\ 
  300 & 0.082 & 0.405 & 0.791 (0.751, 0.832) & 0.051 & 0.383 & 0.814 (0.776, 0.852) \\ 
  330 & 0.082 & 0.428 & 0.791 (0.753, 0.830) & 0.050 & 0.383 & 0.804 (0.765, 0.844) \\ 
  360 & 0.083 & 0.427 & 0.788 (0.747, 0.828) & 0.049 & 0.400 & 0.803 (0.765, 0.842) \\ 
  390 & 0.085 & 0.400 & 0.789 (0.748, 0.830) & 0.050 & 0.383 & 0.810 (0.771, 0.849) \\ 
   \hline
\end{tabular}
\end{adjustbox}
\caption{\label{tab:aucsh44} Overview of high resolution scan results as shown in Figures \ref{fig:h44}. FPR is the observed rate of false positives (for a fixed level $\alpha = 0.05$), FNR is the false negative rate. Area under the curve (AUC) is shown with 95\% confidence intervals. }
\end{table}

@@

In assessing the suitability of the (deterministic) Chumbley Score for
matching striae on bullet lands we have gained valuable insights into
the process: method CS1 as proposed by \citet{hadler} has a strong
dependency on the specific choice of parameters; the defaults suggested
by \citet{hadler} for screw drivers are not directly applicable for the
smaller bullet lands. The coarseness parameter in particular has a
strong impact on the performance of the test. However, we were able to
suggest some heuristics based on the assumption that once sub-class
characteristics are removed, optimal locations are distributed uniformly
across the profile. For bullet lands we found a coarseness value of
\(c = 0.15\) to be suitable for the low-resolution scans from NIST and a
value of \(c=0.125\) suitable for the higher-resolution scans from
CSAFE. Sizes for optimization windows \(w_o\) were based on
cross-validation to minimize overall type 2 error rates. Unfortunately,
this kind of assessment is only feasible in the setting of a large
study, such as the one we presented, and does not transfer immediately
to case work, where a forensic examiner would only deal with a few
identifications. Parameter settings for different studies should be
investigated and reported to allow for a further refinement.

Method CS1 proposed by \citet{hadler} has a minimal type 2 error rate of
0.272 for an optimized window size of 140 pixels -- which is
considerably higher than the error rates achieved on matching toolmarks,
but is similar to other single-feature methods proposed for bullet
matching. Unfortunately, method CS1 also has a high rate of failed tests
-- situations, in which the algorithm does not provide a result, due to
the way different-shift pairs are constructed. Algorithm CS2 is
introduced here as a remedy for failed tests by introducing an alternate
version of choosing different-shift pairs. The algorithm CS2 introduced
here is constructed in a way that achieves on average a ten-fold
reduction in the number of failures. While reducing the failure rate,
the algorithm also shows an increase in the power of the test. Type II
error rates of CS2 reach a minimum of 0.217 for an optimized window size
of 130 pixels. This increase in power of CS2 over CS1 should also apply
to previous studies on toolmarks. It would be interesting to see these
results using the adjusted algorithm. Unfortunately, none of the studies
have made the data publicly accessible.

>>>>

  \author{
        Ganesh Krishnan,\\ %\\
    Master of Science (M.Sc.)\\
    % \\
     and \\     Prof.~Heike Hofmann,\\ %\\
    Doctor of Philosophy (Ph.D.)\\
    % \\
      }
    \maketitle
  \bigskip
  \begin{center}
  {
  \textbf{Affiliation(s)}\\
  Department of Statistics, Iowa State University and CSAFE (Center for
  Statistics and Applications in Forensic Evidence)\\
  \bigskip
  \textbf{Source of funding}\\
  The authors gratefully acknowledge funding from CSAFE through
  cooperative agreement \# 70NANB15H176 between NIST and Iowa State
  University.\\
  \bigskip
  \textbf{Acknowledgement}\\
  We would like to thank the efforts of CSAFE and Roy J Carver Microscopy
  lab in scanning the Hamby-44 set and providing the scans to us. We would
  also like to thank NIST for providing the Hamby-44 bullets to us for
  scanning. We would also like to thank the two anonymous reviewers for
  their helpful suggestions in improving the paper.\\
  \bigskip
  \textbf{Information on presentation of work}\\
  This work was presented in the form of a poster in the AFTE 2018 49th
  Annual Training Seminar, held from June 3rd to June 8th, 2018 in
  Charleston, West Virginia
  }
  \end{center}

@@

Methods for matching striated marks for a variety of tools have been
studied in the literature (see \autoref{tab:toolmarks-ER} for an
overview): in \cite{manytoolmarks1} and \cite{chumbley} digitized
screwdriver marks have been analyzed using a profilometer; in
\cite{manytoolmarks2} 3D marks from screwdriver, tongue and groove
pliers captured using a confocal microscope have been investigated;
digitized marks from slip-joint pliers generated by a surface
profilometer have been investigated in \cite{afte-chumbley}.

These digitized markings allow the use of statistical methods to
quantify the scientific mechanism of comparing markings and serve as
basis for an error rate calculation. In \cite{manytoolmarks2} a relative
distance metric is defined and used as similarity measure between two
toolmarks. This approach is expanded in \citet{manytoolmarks1}: a set of
small segments in the markings of two toolmarks are extracted and
similarity is compared using a maximum Pearson correlation coefficient.
The Chumbley scoring method, first introduced in \cite{chumbley}, uses a
similar but more extensive framework based on a Mann-Whitney U test of
the resulting correlation coefficients. This approach is
non-deterministic, because segments are chosen randomly. In
\cite{hadler} the score is fixed to be deterministic for each pair of
toolmarks by choosing segments for comparison systematically. This
approach also ensures independence between segments of striae.

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{\begin{tabular}{rlrrr}
  \hline
Research paper & Method & Data Source & False Positives & False Negatives\\
  \hline
 \textbf{Faden et al. (2007)} \cite{manytoolmarks1} & Maximum Pearson & Screwdrivers & & \\ & Correlation &  & - & - \\ \hline
 \textbf{Chumbley et al. (2010)} \cite{chumbley} & Randomized & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 2.3\% & 8.9\%  \\ \hline
  \textbf{Grieve et al. (2014)}
 \cite{afte-chumbley} & Randomized & Slip-joint  &  & \\ & Chumbley Score &  & - & - \\ \hline 
  \textbf{Hadler \& Morris (2017)} \cite{hadler} & Deterministic & Screwdrivers & & \\ (Same-Surface Same-Angle) & Chumbley Score & & 0\% & 6\%  \\ \hline
  \textbf{Bachrach et al. (2010)} \cite{manytoolmarks2} & Similarity Measure & Screwdrivers & & \\ (Different Surfaces-same angle) & Relative Distance Metric & & 5.9\% &  9.4\% \\ (Same Surfaces-same angle)&  & & 0.22\% &  0\% \\
   \hline
\end{tabular}}
\caption{\label{tab:toolmarks-ER} Error Rates in same-source analyses for  striated toolmarks reported in the literature. The top four reported papers use some variation of the Chumbley score method.}
\end{table}

@@

\begin{table}[hbtp]
\centering
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{rrrlrrl}
  \hline
  & \multicolumn{3}{l}{\bf CS1} & \multicolumn{3}{l}{\bf CS2} \\
$w_o$ & \bf \%FP & \bf \%FN & \bf AUC (95\% C.I.) & \bf \%FP & \bf \%FN & \bf AUC (95\% C.I.) \\ 
  \hline
90 & 6.8 & 33.0 & 0.850 (0.835, 0.865) & 7.5 & 24.2 & 0.877 (0.863, 0.890) \\ 
  120 & 6.5 & 30.9 & 0.864 (0.850, 0.878) & 7.2 & 23.6 & 0.890 (0.877, 0.903) \\ 
  130 & 6.5 & 31.7 & 0.863 (0.850, 0.877) & 7.2 & 22.4 & 0.898 (0.886, 0.910) \\ 
  140 & 6.5 & 30.4 & 0.873 (0.859, 0.886) & 7.1 & 22.9 & 0.902 (0.891, 0.914) \\ 
  150 & 6.7 & 32.6 & 0.863 (0.850, 0.877) & 7.2 & 22.5 & 0.904 (0.892, 0.916) \\ 
  180 & 6.1 & 32.6 & 0.877 (0.864, 0.890) & 7.0 & 22.9 & 0.907 (0.896, 0.919) \\ 
  210 & 6.4 & 34.2 & 0.865 (0.851, 0.878) & 6.6 & 23.1 & 0.906 (0.895, 0.918) \\    \hline
\end{tabular}
\end{adjustbox}
\caption{\label{tab:aucs} Overview of results as shown in Figures \ref{fig:roc} and \ref{fig:eer}. \%FP is the observed percentage of false positives (for a fixed level $\alpha = 0.05$), \%FN is the percentage of false negatives. Area under the curve (AUC) is shown with confidence intervals. }
\end{table}

@@

\autoref{fig:type1} shows the percentages of observed type I errors
(\%FP) across a range of optimization windows \(w_o\). Generally,
observed type I errors are higher than expected. Method CS1 shows in
this instance slightly better performance than method CS2, but for both
an increase in the size of the optimization window leads to a decrease
in the observed type I errors.

@@

The high-resolution scans of Hamby set 44 are capturing images at a
resolution of \(0.645 \mu m\) per pixel. On average, land engraved areas
are 3000 pixels in length. A coarseness of \(c = 0.125\) seemed to be
sufficient in removing any bullet curvature. Both methods have a failed
test rate of less than 0.006, indicating, again, that the larger number
of pixels alleviates the problem of test failures. \autoref{fig:h44}
shows the resulting EER and AUC for methods CS1 and CS2 based on two
sizes of validation windows (\(w_v \in 75, 125\)) and optimization
window sizes around 300 pixels (10 percent of the average length), with
corresponding errors shown numerically in \autoref{tab:aucsh44}. Both
methods show an increase in performance around \(w_o = 300\) pixels. CS2
out-performs CS1 in all scenarios, but the difference is not significant
(using DeLong's confidence intervals). Interestingly, the overall
performance of both CS1 and CS2 is a lot lower for the high resolution
version of Hamby-44 than for the lower resolution scans. The area under
the curve overall is significantly lower for the high resolution scans
than for the previous set of scans. Partly, this might be due to the
particular choice of the parameters, partly the higher-resolution scans
might be picking up on real differences between the lands that the
lower-resolution scans fail to detect. Re-assuringly, the observed error
rate of false positives for CS2 is closer to the nominal rate of 5\%
than for the lower-resolution scans.

@@

\begin{table}[hbtp]
\centering
\begin{adjustbox}{max width=\textwidth}
\begin{tabular}{rrrlrrl}
  \hline
  & \multicolumn{3}{l}{\bf CS1} & \multicolumn{3}{l}{\bf CS2} \\
$w_o$ & \bf \%FP & \bf \%FN & \bf AUC (95\% C.I.) & \bf \%FP & \bf \%FN & \bf AUC (95\% C.I.) \\  
  \hline
  210 & 8.6 & 41.3 & 0.780 (0.740, 0.821) & 5.5 & 38.7 & 0.803 (0.763, 0.843) \\ 
  240 & 8.5 & 42.7 & 0.784 (0.744, 0.824) & 5.5 & 36.5 & 0.801 (0.760, 0.841) \\ 
  270 & 8.2 & 41.5 & 0.782 (0.741, 0.823) & 4.8 & 37.4 & 0.808 (0.769, 0.848) \\ 
  300 & 8.2 & 40.5 & 0.791 (0.751, 0.832) & 5.1 & 38.3 & 0.814 (0.776, 0.852) \\ 
  330 & 8.2 & 42.8 & 0.791 (0.753, 0.830) & 5.0 & 38.3 & 0.804 (0.765, 0.844) \\ 
  360 & 8.3 & 42.7 & 0.788 (0.747, 0.828) & 4.9 & 40.0 & 0.803 (0.765, 0.842) \\
  390 & 8.5 & 40.0 & 0.789 (0.748, 0.830) & 5.0 & 38.3 & 0.810 (0.771, 0.849) \\ 
   \hline
\end{tabular}
\end{adjustbox}
\caption{\label{tab:aucsh44} Overview of high resolution scan results as shown in Figures \ref{fig:h44}. \%FP is the percentage of observed  false positives (for a fixed level $\alpha = 0.05$), \%FN is the percentage of false negatives. Area under the curve (AUC) is shown with 95\% confidence intervals. }
\end{table}

@@

In assessing the suitability of the (deterministic) Chumbley Score for
matching striae on bullet lands we have gained valuable insights into
the process: method CS1 as proposed by \citet{hadler} has a strong
dependency on the specific choice of parameters; the defaults suggested
by \citet{hadler} for screw drivers are not directly applicable for the
smaller bullet lands. The coarseness parameter in particular has a
strong impact on the performance of the test. However, we were able to
suggest some heuristics based on the assumption that once sub-class
characteristics are removed, optimal locations are distributed uniformly
across the profile. For bullet lands we found a coarseness value of
\(c = 0.15\) to be suitable for the low-resolution scans from NIST and a
value of \(c=0.125\) suitable for the higher-resolution scans from
CSAFE. Sizes for optimization windows \(w_o\) were based on
cross-validation to minimize overall type 2 error rates. Ideally, the
exact values for parameters will be determined in a large study
incorporating different types of firearms and brands of ammunition.
\%Unfortunately, this kind of assessment is only feasible in the setting
of a large study, such as the one we presented, Results therefore do not
transfer immediately to case work, where a forensic examiner would only
deal with a few identifications. \%Parameter settings for different
studies should be investigated and reported to allow for a further
refinement.

Method CS1 proposed by \citet{hadler} has a minimal type 2 error rate of
0.272 for an optimized window size of 140 pixels -- which is
considerably higher than the error rates achieved on matching toolmarks,
but is similar to other single-feature methods proposed for bullet
matching. Unfortunately, method CS1 also has a high rate of failed tests
-- situations, in which the algorithm does not provide a result, due to
the way different-shift pairs are constructed. Algorithm CS2 is
introduced here as a remedy for failed tests by introducing an alternate
version of choosing different-shift pairs. Algorithm CS2 is constructed
in a way that achieves on average a ten-fold reduction in the number of
failures. While reducing the failure rate, the algorithm also shows an
increase in the power of the test. Type II error rates of CS2 reach a
minimum of 0.217 for an optimized window size of 130 pixels. This
increase in power of CS2 over CS1 should also apply to previous studies
on toolmarks. It would be interesting to see these results using the
adjusted algorithm. Unfortunately, none of the studies have made the
data publicly accessible.
